### 2016-05-11 main.tex
Unique structural derivation and edits (Chapter 6)
Add structural coeffect proofs (Chapter 6)
Switch order of chapters 7 and 8
Add binding to chapter 2, whitespace, tweaks

% Includes the file which contains all the document configurations and packages - make sure to edit this file
\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{catchfilebetweentags}

@@

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\ttfamily #1}}}
\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\bfseries\sffamily #1}}}
\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

@@

@@

% For definining semantics of things over a typing derivation
\newcommand{\semdef}[2]{
\begin{array}{rcl}
  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
  \sem{#1} & = & #2 \\
\end{array}
}
\newcommand{\semdeff}[4]{
\begin{array}{rcl}
  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
  \sem{#2} & = & #4 \\
\end{array}
}
\newcommand{\emdeff}[4]{
\begin{array}{rcl}
  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
  #2 & = & #4 \\
\end{array}
}
\newcommand{\semdefff}[6]{
\begin{array}{rcl}
  \sem{#1} & = & #4 \\
  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
  \sem{#3} & = & #6 \\
\end{array}
}
\newcommand{\emdefff}[6]{
\begin{array}{rcl}
  \sem{#1} & = & #4 \\
  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
  #3 & = & #6 \\
\end{array}
}

@@

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

% \ctparttext{You can put some informational part preamble text here. Illo principalmente su nos. Non message \emph{occidental} angloromanic da. Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via. Pan in terra summarios, capital interlingua se que. Al via multo esser specimen, campo responder que da. Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 1 page describing  the content in Part 1

% \part{Some Kind of Manual} % First part of the thesis

% \include{Chapters/Chapter01} % Chapter 1

% \cleardoublepage % Empty page before the start of the next part

%------------------------------------------------

% \ctparttext{You can put some informational part preamble text here. Illo principalmente su nos. Non message \emph{occidental} angloromanic da. Debitas effortio simplificate sia se, auxiliar summarios da que, se avantiate publicationes via. Pan in terra summarios, capital interlingua se que. Al via multo esser specimen, campo responder que da. Le usate medical addresses pro, europa origine sanctificate nos se.} % Text on the Part 2 page describing the content in Part 2

% \part{The Showcase} % Second part of the thesis

%\include{text/Chapter01}
%\include{text/Chapter02}
%\include{text/Chapter03}

@@

  The following page provides a detailed overview of the content of Chapters~\ref{ch:flat} and
  Chapters~\ref{ch:semantics}, highlighting the split between general definitions and properties
  (about the coeffect calculus) and concrete definitions and properties (about concrete
  context-awre language). The Chapter~\ref{ch:structural} mirrors the same development for
  \emph{structural coeffect systems}.
}
\setcounter{part}{1}
\part{Coeffect calculi}
\label{part:coeffect-calculi}

@@

\include{text/Chapter04}
\include{text/Chapter05}
%\include{text/Chapter06}
%\include{text/Chapter07}
\include{text/Chapter08}
%\include{text/Chapter09}

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - APPENDICES
%----------------------------------------------------------------------------------------

% \part{Appendix} % New part of the thesis for the appendix

% \include{Chapters/Chapter0B} % Appendix B - empty template

%----------------------------------------------------------------------------------------
%	POST-CONTENT THESIS PAGES
%----------------------------------------------------------------------------------------

\cleardoublepage\include{Bibliography} % Bibliography
% \cleardoublepage\include{FrontBackMatter/Colophon} % Colophon
% \cleardoublepage\include{FrontBackMatter/Declaration} % Declaration

%----------------------------------------------------------------------------------------

>>>>

% Includes the file which contains all the document configurations and packages - make sure to edit this file
\input{classicthesis-config}
\usepackage{semantic}
\usepackage{paralist}
\usepackage{etoolbox}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumitem}
%\usepackage{catchfilebetweentags}

@@

% Formatting for source code & types
\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.3,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\ttfamily #1}}}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\prepk}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\prepi}[1]{\textnormal{\textcolor{prepclr}{\ttfamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

@@

\urlstyle{sf}

@@

% For definining semantics of things over a typing derivation
\newcommand{\semdef}[2]{
\begin{array}{rcl}
  ~ & ~ & ~ \\ \cline{1-1}\cline{3-3}
  \sem{#1} & = & #2 \\
\end{array}
}
\newcommand{\semdeff}[4]{
\begin{array}{rcl}
  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
  \sem{#2} & = & #4 \\
\end{array}
}
\newcommand{\mdeff}[4]{
\begin{array}{rcl}
  #1 & = & #3 \\ \cline{1-1}\cline{3-3}
  #2 & = & #4 \\
\end{array}
}
\newcommand{\emdeff}[4]{
\begin{array}{rcl}
  \sem{#1} & = & #3 \\ \cline{1-1}\cline{3-3}
  #2 & = & #4 \\
\end{array}
}
\newcommand{\semdefff}[6]{
\begin{array}{rcl}
  \sem{#1} & = & #4 \\
  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
  \sem{#3} & = & #6 \\
\end{array}
}
\newcommand{\emdefff}[6]{
\begin{array}{rcl}
  \sem{#1} & = & #4 \\
  \sem{#2} & = & #5 \\ \cline{1-1}\cline{3-3}
  #3 & = & #6 \\
\end{array}
}

@@

\ctparttext{  The computing ecosystem is becoming increasingly heterogeneous and rich. Modern
  programs need to run on a variety of devices that are all different, but provide unique rich
  capabilities. For example, application running on a phone can access the GPS sensor, while
  application running in the cloud can access GPU computing resources. Both diversity and richness
  can only be expected to increase with trends such as the internet of things.

  In this thesis, we argue that the creating programming languages that allow the programmer to
  better work with the environment or \emph{context} in which applications execute is the next
  big challenge for programming language designers.

  We start with a detailed discussion of the motivation for the thesis and an overview of our
  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming langauge research that
  leads to the work presented in this thesis (Chapter~\ref{ch:pathways}) and we examine a number
  of practical context-aware systems in detail (Chapter~\ref{ch:applications}), identifying two
  kinds of context that we later capture by \emph{flat} and \emph{structural coeffects}.
}
\part{Context-aware programming}
\label{part:context-aware}

\include{text/Chapter01}
\include{text/Chapter02}
\include{text/Chapter03}

@@

  The following page provides a detailed overview of the content of Chapters~\ref{ch:flat} and
  Chapters~\ref{ch:semantics}, highlighting the split between general definitions and properties
  (about the coeffect calculus) and concrete definitions and properties (about concrete
  context-awre language). The Chapter~\ref{ch:structural} mirrors the same development for
  \emph{structural coeffect systems}.
}
\part{Coeffect calculi}
\label{part:coeffect-calculi}

% --------------------------------------------------------------------------------------------------

@@

%\include{text/Chapter04}
%\include{text/Chapter05}
%\include{text/Chapter06}

\ctparttext{ TBD }
\part{Towards practical coeffects}
\label{part:context-aware}

%\include{text/Chapter07}
%\include{text/Chapter08}
%\include{text/Chapter09}

\cleardoublepage
\include{Bibliography}

% --------------------------------------------------------------------------------------------------

### 2016-05-11 text/Chapter01.tex
Further work and conclusions (Chapter 7)
Add binding to chapter 2, whitespace, tweaks

%!TEX root = ../main.tex

% ==================================================================================================

@@

Putting deeper philosophical questions about the nature of scientific prog\-ress aside,
the goal of programming language research is generally to design languages that provide
more \emph{appropriate abstractions} for capturing common problems, are \emph{simple} and
more \emph{unified}. These are exactly the aims that we follow in this thesis. In this chapter, 
we explain what the common problems in context-dependent programming are. In 
Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we develop two simple calculi to 
understand and capture the structure of those problems and, finally, Chapter~\ref{ch:unified} 
unifies the two abstractions.

% ==================================================================================================
%                                                                                             
%     ###                  #                    #                                              
%    #   #                 #                    #                                              
%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###  
%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   # 
%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      ##### 
%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #     
%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###  
%                                                                                                                                                                       % ==================================================================================================

\section{Why context-aware programming matters}

@@

\item \textsc{Security and provenance.} When working with data (be it sensitive database or social
   network data), we may have permission to access only some of the data and we may want to track
   \emph{provenance} information. However, this is not checked -- if a program attempts to access
   unavailable data, the access will be refused at run-time.

@@

For example, the following LINQ query written in C\# queries a database and
selects those product names where the first upper case letter is ``C'':
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

@@

For example, consider the following code sample that accesses a database by building a SQL query
using string concatenation. For the purpose of the demonstration, this example does not use LINQ, 
but an older approach with a parameterized SQL query written as a string:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as their 
\ident{name} and delete the database table ``Products''! For this reason, most libraries discourage 
building SQL commands by string concatenation, but there are still many systems that do so.

Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
meta-data about variables that are in some ways special. Such meta-data can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

In the security context, such marking of values (but at run-time) is called \emph{tainting} 
\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance}
tracking \cite{app-provenance-future}. This can be useful when working with data in other contexts. 
For example, data jounralsts might want to propagate meta-data about the quality and the 
information source -- is the source trustworthy? Is the data up-to-date? Such meta-data could 
propagate to the result and tell us important information about the calculated results.

@@

The information about access patterns can be used to efficiently compile the computation
by preallocating the necessary space (as we know which sub-range of the array might be accessed).
It also allows better handling of boundaries \cite{app-stencil-dsl}. For example, to simplify 
wrap-around behaviour we could pad the \ident{input} with a known number of elements from the 
other side of the array.

@@

The four examples presented in the previous section cover different kinds of \emph{context}.
The context includes notions such as execution environment, capabilities provided by the 
environment or input and meta-data about the input and variables through which it is accessed.

@@

\begin{itemize}
\item \textsc{Flat coeffects} -- represent additional data, resources and meta-data that are
   available in the execution environment (regardless of how they are accessed in a program).
   Examples include resources such as GPS sensors and battery status (on a phone), databases
   (on the server), or software framework (or library) version.

\item \textsc{Structural coeffects} -- capture additional meta-data related to inputs. This can include
   provenance (source of the input value), usage information (how often is the value
   accessed and in what ways) or security information (whether it contain sensitive data or not).
\end{itemize}
%
This thesis follows the tradition of statically typed programming languages. As such, we
attempt to capture such contextual information in the type system of context-aware programming
languages. The type system should provide both safety guarantees (as in the first three examples)
and also static analysis useful for optimization (as in the last example).

@@

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing mouse
pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{\textbf{fetchLocalNews}}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

The details of how the cross-compilation would work are out of the scope of this thesis. However,
one can imagine that the compiler would take multiple sets of references (representing the different
platforms), expose the \emph{union} of the functions, but annotate each with the required platform.
Then, it would produce multiple different binaries -- here, one for the server-side (containing
\ident{fetchNews}), one for iPhone and one for Windows.

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float}, or even \ident{string}, depending on the types of the variables. For example, 
the following is a valid typing judgement in F\# \cite{app-fsharp}:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.

@@

The systems for tracking of context requirements developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this 
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks meta-data about the variables and annotates \ident{password} 
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

\begin{itemize}
\item In Chapter~\ref{ch:pathways}, we give an overview of related work. Most importantly, we show
  that the idea of context-aware computations can be naturally approached from a number of
  directions developed recently in the theory of programming languages (including type and effect
  systems, categorical semantics and sub-structural logics).

@@

\item In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we present the key
  contributions of this thesis. We develop the \emph{flat} and \emph{structural} calculi, show
  how they capture important contextual properties and develop their categorical semantics using
  a notion based on comonads. Chapter~\ref{ch:unified}  links the two systems using a single 
  formalism that is capable of capturing both flat and structural properties and also discusses
  an alternative presentation that is more suitable for automatic type inference of coeffects.

\item Related work is presented in Chapter~\ref{ch:pathways} and throughout the thesis. 
  One important direction deserves further exploration, and so Chapter~\ref{ch:unified}
  also includes a brief discussion of a different approach to tracking contextual information that 
  arises from modal logics. This is followed by an outline of the future work that is needed
  in order to implement the presented theory in main-stream programming languages.
\end{itemize}

>>>>

@@

% ==================================================================================================
%
%     ###                  #                    #
%    #   #                 #                    #
%    #       ###   # ##   ####    ###   #   #  ####           ###   #   #   ###   # ##    ###
%    #      #   #  ##  #   #     #   #   # #    #     #####      #  #   #      #  ##  #  #   #
%    #      #   #  #   #   #     #####    #     #             ####  # # #   ####  #      #####
%    #   #  #   #  #   #   #  #  #       # #    #  #         #   #  # # #  #   #  #      #
%     ###    ###   #   #    ##    ###   #   #    ##           ####   # #    ####  #       ###
%                                                                                                                                                                       % ==================================================================================================
% ==================================================================================================

\section{Why context-aware programming matters}
\label{sec:intro-whymatters}

@@

\item \textsc{Security and provenance.} When working with data (be it sensitive database or social
   network data), we may have permission to access only some of the data and we may want to track
   \emph{provenance} information. However, this is not checked statically -- if a program attempts
   to access unavailable data, the access will be refused at run-time.

@@

For example, the following LINQ query written in C\# queries a database and
selects those product names where the first upper case letter is ``C'':
%
\begin{equation*}
\begin{array}{l}
\kvd{var}~\ident{db} = \kvd{new}~\ident{NorthwindDataContext}();\\[0.25em]
\kvd{from}~\ident{p}~\kvd{in}~\ident{db.Products}\\[-0.25em]
\kvd{where}~\ident{p.ProductName.First}(\lambda \ident{c} \rightarrow \ident{Char.IsUpper}(\ident{c})) == \str{C}\\[-0.25em]
\kvd{select}~\ident{p.ProductName};
\end{array}
\end{equation*}
%
This appears as a perfectly valid code and the C\# compiler accepts it. However, when the program
is executed, it fails with the following error:
%
\begin{quote}
\emph{Unhandled Exception: \texttt{System.NotSupportedException}: Sequence operators
not supported for type \texttt{System.String}.}
\end{quote}
%
The problem is that LINQ can only translate a \emph{subset} of normal C\# code. The above snippet
uses the \ident{First} method to iterate over characters of a string, which is not supported. This is
not a technical limitation of LINQ, but a fundamental problem of the approach.

@@

For example, consider the following code sample that accesses a database by building a SQL query
using string concatenation. For the purpose of the demonstration, this example does not use LINQ,
but an older approach with a parameterized SQL query written as a string:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{query} = \ident{sprintf}~\str{SELECT * FROM Products WHERE Name='\%s'}~\ident{name}\\[-0.25em]
\kvd{let}~\ident{cmd} = \kvd{new}~\ident{SqlCommand}(\ident{query})\\[-0.25em]
\kvd{let}~\ident{reader} = \ident{cmd.ExecuteReader}()
\end{array}
\end{equation*}
%
The code compiles without error, but it contains a major security flaw called \emph{SQL injection}
\cite{app-sql-injection}. An attacker could enter \str{'; DROP TABLE Products -{}-} as the
\ident{name} and delete the database table ``Products''. For this reason, most libraries discourage
building SQL commands by string concatenation, but there are still many systems that do so.

Again, this example demonstrates a more general property. Sometimes, it is desirable to track additional
metadata about variables that are in some ways special. Such metadata can determine how the variables
can be used. Here, \ident{name} comes from the user input. This information about the value should
be propagated to \ident{query}. The \ident{SqlCommand} object should then require arguments that can
not directly contain user input (in an unchecked form).

@@

In the security context, such marking of values (but at run-time) is called \emph{tainting}
\cite{app-tainting-sql}, but the technique is a special case of more general \emph{provenance
tracking} \cite{app-provenance-future}. This can be useful when working with data in other contexts.
For example, data journalists might want to propagate metadata about the quality and the
information source -- is the source trustworthy? Is the data up-to-date? Such metadata could
propagate to the result and tell us important information about the calculated results.

@@

Access patterns can be used to efficiently compile the computation
by preallocating the necessary space (as we know which sub-range of the array might be accessed).
It also allows better handling of boundaries \cite{app-stencil-dsl}. For example, to simplify
wrap-around behaviour we could pad the \ident{input} with a known number of elements from the
other side of the array.

@@

The four examples presented in the previous section cover different kinds of \emph{context}.
The context includes notions such as execution environment, capabilities provided by the
environment or input and metadata about the input and variables through which it is accessed.

@@

\begin{itemize}
\item \textsc{Flat coeffects} represent additional data, resources and metadata that are
   available in the execution environment (regardless of how they are accessed in a program).
   Examples include resources such as GPS sensors and battery status (on a phone), databases
   (on the server), or software framework (or library) version.

\item \textsc{Structural coeffects} capture additional metadata related to inputs. This can include
   provenance (source of the input value), usage information (how often is the value
   accessed and in what ways) or security information (whether it contain sensitive data or not).
\end{itemize}
%
This thesis follows the tradition of statically typed programming languages. As such, we
attempt to capture such contextual information in the type system of context-aware programming
languages. The type system should provide both safety guarantees (as in the first three examples)
and also static analysis useful for optimization (as in the last example).

@@

Furthermore, integrating contextual information into the type system can provide information for
modern developer tools. For example, many editors for F\# display inferred types when placing mouse
pointer over an identifier. For \ident{fetchLocalNews}, the tip could appear as follows:
%
\definecolor{tooltipy}{rgb}{0.98,0.97,0.76}
\definecolor{tooltipb}{rgb}{0.7,0.7,0.7}
\vspace{0.5em}
\begin{equation*}
\hspace{5em}
\begin{array}{l}
\ident{fetchLocalNews}\\
\hspace{1em} \fcolorbox{tooltipb}{tooltipy}{
	\introfuncblack{unit}{gps, rpc}{(news~list)~async}
}
\end{array}
\end{equation*}
\vspace{0.5em}

@@

\noindent
The details of how the cross-compilation would work are out of the scope of this thesis. However,
one can imagine that the compiler would take multiple sets of references (representing the different
platforms), expose the \emph{union} of the functions, but annotate each with the required platform.
Then, it would produce multiple different binaries -- here, one for the server-side (containing
\ident{fetchNews}), one for iPhone and one for Windows.

@@

\paragraph{Type systems.}
A type system is a form of static analysis that is usually specified by \emph{typing judgements}
such as $\Gamma \vdash e : \tau$. The judgement specifies that, given some variables described by
the context $\Gamma$, the expression $e$ has a type $\tau$.
The variable context $\Gamma$ is necessary to determine the type of expressions. Consider an
expression $x + y$. In many languages, including Java, C\# and F\#, the type could be \ident{int},
\ident{float} or \ident{string}, depending on the types of the variables. For example,
the following is a valid typing judgement in F\# \cite{app-fsharp}:
%
\begin{equation*}
x\!:\!\ident{int},~y\!:\!\ident{int} \vdash x+y : \ident{int}
\end{equation*}
%
This judgement assumes that the type of both $x$ and $y$ is \ident{int} and so the result must also
be \ident{int}. In F\#, the expression would also be typeable in a context $x\!:\!\ident{string},~y\!:\!\ident{string}$,
but not, for example, in a context where $x$ has a type \ident{int} and $y$ has a type \ident{string}.

@@

The systems for tracking of context requirements developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks metadata about the variables and annotates \ident{password}
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

\begin{itemize}
\item In Chapter~\ref{ch:pathways}, we give an overview of related work. Most importantly, we show
  that the idea of context-aware computations can be naturally approached from a number of
  directions developed recently in the theory of programming languages (including type and effect
  systems, categorical semantics and substructural logics).

@@

\item In Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}, we capture important contextual
  properties using a simple theoretical models. We develop the \emph{flat coeffect calculus} that
  captures per-context properties and \emph{structural coeffect calculus} that captures
  per-variable properties. We give a type system for the calculi and study their equational
  properties.

\item In Chapter~\ref{ch:semantics} we give categorical semantics of the flat coeffect calculus.
  This provides a unified way of defining the semantics of context-aware languages. We use the
  categorical semantics as a basis for \emph{categorically-inspired translation} that turns
  context-aware programs into programs in a simple functional language. We prove that well-typed
  context-aware programs are translated to programs that ``do not go wrong''. The development
  is repeated for structural coeffects in Chapter~\ref{ch:structural}.

\item In Chapter~\ref{ch:impl}, we use the translation as a basis for a prototype
  implementation of three simple context-aware programming languages. Our implementation serves as
  an empirical evidence showing how coeffects simplify programming with context and provide
  additional safety guarantees. We present the implementation in the format of an \emph{interactive
  essay} (\url{http://tomasp.net/coeffects}) that encourages active reading and exploration of
  various aspects of the theory.

\item Related work is presented in Chapter~\ref{ch:pathways} and, together with further work,
  throughout the thesis. Two important directions deserve furhter exploration. In Chapter~\ref{ch:further},
  we outline a unified coeffect system that is capable of capturing both flat and structural
  properties. We also include a brief discussion of a different approach to tracking contextual
  information that arises from modal logics.
\end{itemize}
